AFQ-Insight

### 2019-12-06 reports/manuscript/sections/conclusion.tex

The results we present here also motivate extensions of the method using more
sophisticated cost functions. For example, the fused sparse group lasso (FSGL)
\cite{zhou2012} extends SGL to enforce additional spatial structure: smoothness
in the variation of diffusion metrics along the tracts. As brain measurements
include additional structure (for example, bilateral symmetry), future work
could also incorporate overlapping group membership for each entry in the tract
profiles \ariel{citation needed}. For example, a measurement could come from the
corpus callosum, but also from the right hemipshere. This would also require
extending the cost function used here to incorporate these constraints.

>>>>

The results we present here also motivate extensions of the method using more
sophisticated cost functions. For example, the fused sparse group lasso (FSGL)
\cite{zhou2012} extends SGL to enforce additional spatial structure: smoothness
in the variation of diffusion metrics along the tracts. As brain measurements
include additional structure (for example, bilateral symmetry), future work
could also incorporate overlapping group membership for each entry in the tract
profiles \cite{Rao2014-xm}. For example, a measurement could come from the
corpus callosum, but also from the right hemipshere. This would also require
extending the cost function used here to incorporate these constraints.

### 2019-12-06 reports/manuscript/sections/introduction.tex

To relate the diffusion in each voxel to the macro-structure of long-range
connections between different brain regions, methods for computational
tract-tracing from diffusion MRI, or tractography, combine the estimates of
fiber orientations in each voxel to form streamlines that traverse the volume of
the white matter \cite{Conturo1999-je, Mori2002-qi}. These methods have been
under increased scrutiny and several lines of investigation have raised
critiques of their validity \cite{Maier-Hein2017-vb, Thomas2014-ki}. On the
other hand, there have been efforts to shore up the inferences made with these
methods \cite{Pestilli2014NatMeth, Takemura2016-sh, Smith2013-nc, Smith2015-cx,
Smith2015-zt, Rheault2018-wk}. Importantly, though discovering novel fiber
bundles requires extraordinary evidence, and delineating the exact location of
the termination of the streamlines in the gray matter, there is little dispute
that tractography can accurately delineate major white matter bundles that are
known to exist within the core of the white matter \cite{Maier-Hein2017-vb}.

Using this fact, one of the most powerful methods currently available to put
macro- and micro-structure together is \emph{tractometry} (also known as
tract-based morphometry): assessment of the physical properties of the white
matter along specific tracts \cite{Bells2011-cf}. Though there are several
different available implementations of this overall idea, the principles are
similar \cite{yeatman2012tract, Yendiki2011-ay, Wassermann2016-iv, ODonnell2009-uu}:
tractometry begins by delineating the parts of the white matter that belong to
different major tracts and axonal fiber bundles, such as the corticospinal tract
or arcuate fasciculus, and samples the biophysical properties (such as
fractional anisotropy or mean diffusivity) along the length of these bundles. In
some cases, tissue properties along the length of each bundle are summarized by
taking the mean along each bundle, but there is a large body of evidence showing
that there is systematic variability within each bundle. This justifies
retaining the individual samples along the length of each fiber
\cite{yeatman2012tract, colby2012, ODonnell2009-uu}. However, while this retains
important information about each individual's white matter, it also presents
statistical challenges due to the dimensionality of the data. Based on
tractometry, researchers may choose to compare different individuals to each
other. This is usually done following one of the following approaches:

@@

\item ROI-based selection, followed by multivariate analysis. Here, an ROI is
selected based on \emph{a priori} knowledge, and all the nodes or voxels in the
ROI are used together to fit a model that can predict differences between
individuals. An example of that is the ``profilometry'' framework, in which
different diffusion metrics from a single tract are combined together to provide
input to a multivariate analysis of covariance, and linear discriminant analysis
\cite{dayan2016profilometry}. Other work has used random forests algorithms
applied to the corticospinal tract to discriminate between patients with ALS and
matched controls \cite{sarica2017corticospinal}

@@

In the present work, we present a novel framework for analysis of tractometry
that simultaneously selects the features for analysis, and fits a model to these
features. Rather than a comparative approach, this framework takes a predictive
approach to inference \cite{Breiman2001-uz}. We use a linear modeling approach,
which aims to predict phenotypical variance in a group of subjects, based on a
linear combination of the features estimated with tractometry. Using this
approach, we need to first deal with the large and asymmetric dimensionality of
the data: tractometry data usually has many more features (i.e., number of
measurements per individual) than samples (number of subjects), which makes
inferences from the data about phenotypical differences between individuals
ill-posed. This regime is the target of several statistical learning techniques,
and is often solved by various forms of regularization. For example, Tikhonov
regularization shrinks the solution such that the sum of squared contributions
from the individual features are minimized \cite{Hoerl2000-ij}. Another solution
to the problem is provided by the Lasso algorithm, which instead minimizes the
sum of the absolute values of contributions of each feature
\cite{Tibshirani1996-qs}. This tends to shrink to zero the contributions of many
of the features, providing results that are both accurate and interpretable.
When additional structure is available in the organization of the data,
regularization algorithms can take advantage of this structure. For example, if
the features lend themselves to a natural division into different groups, the
group lasso (GL) can be used to select groups of features, rather than
individual features \cite{Yuan2006-ky}. The Sparse Group Lasso (SGL) elaborates
on this idea by providing control both of group sparsity, as well as overall
sparsity of the solutions \cite{simon2013sgl}. Because the features measured
with tractomery lend themselves to grouping based on the tracts from which each
measurement is taken, GL and SGL provide a useful tool linear model fitting in
problems of this form. Using linear regression as the basis for analysis is
useful for both regression tasks, in which a continuous measure (e.g., age) is
modeled, as well as for classification tasks, where a discrete property is
modeled (e.g., whether an individual has a particular disease). Moreover, use of
SGL in particular helps derive informative properties of the human white matter,
both in cases where the underlying causes of variance are sparsely concentrated
(e.g., in one particular white matter tract) and in cases where biological
substrates of differences between individuals are more widespread. This is
because SGL is controlled by meta-parameters that set the degree to  which and
that can be set empirically, based on the data, in a process of nested
cross-validation. We demonstrate these properties of the algorithm in data from
previously published studies \cite{sarica2017corticospinal,
yeatman2014lifespan}.

>>>>

To relate the diffusion in each voxel to the macro-structure of long-range
connections between different brain regions, methods for computational
tract-tracing from diffusion MRI, or tractography, combine the estimates of
fiber orientations in each voxel to form streamlines that traverse the volume of
the white matter \cite{Conturo1999-je, Mori2002-qi}. These methods have been
under increased scrutiny and several lines of investigation have raised
critiques of their validity \cite{Maier-Hein2017-vb, Thomas2014-ki}. On the
other hand, there have been efforts to shore up the inferences made with these
methods \cite{Pestilli2014NatMeth, Takemura2016-sh, Smith2013-nc, Smith2015-cx,
Smith2015-zt, Rheault2018-wk}. Importantly, though discovering novel fiber
bundles requires extraordinary evidence, and delineating the exact location of
the termination of the streamlines in the gray matter is still prone to error,
there is little dispute that tractography can accurately delineate major white
matter bundles that are known to exist within the core of the white matter
\cite{Maier-Hein2017-vb}.

Using this fact, one of the most powerful methods currently available to put
macro- and micro-structure together is \emph{tractometry} (also known as
tract-based morphometry): assessment of the physical properties of the white
matter along specific tracts \cite{Bells2011-cf}. Though there are several
different available implementations of this overall idea, the principles are
similar \cite{yeatman2012tract, Yendiki2011-ay, Wassermann2016-iv, ODonnell2009-uu}:
tractometry begins by delineating the parts of the white matter that belong to
different major tracts and axonal fiber bundles, such as the corticospinal tract
or arcuate fasciculus, and samples the biophysical properties (such as
fractional anisotropy or mean diffusivity) along the length of these bundles. In
some cases, tissue properties along the length of each bundle are summarized by
taking the mean along each bundle, but there is a large body of evidence showing
that there is systematic variability within each bundle. This justifies
retaining the individual samples along the length of each fiber
\cite{yeatman2012tract, colby2012, ODonnell2009-uu}. While this retains
important information about each individual's white matter, it also presents
statistical challenges due to the dimensionality of the data. Based on
tractometry, researchers may choose to compare different individuals to each
other. This is usually done following one of the following approaches:

@@

\item ROI-based selection, followed by multivariate analysis. Here, an ROI is
selected based on \emph{a priori} knowledge, and all the nodes or voxels in the
ROI are used together to fit a model that can predict differences between
individuals. An example of that is the ``profilometry'' framework, in which
different diffusion metrics from a single tract are combined together to provide
input to a multivariate analysis of covariance, and linear discriminant analysis
\cite{dayan2016profilometry}.

@@

In the present work, we developed a novel framework for analysis of tractometry
that simultaneously selects the features for analysis, and fits a model to these
features. Rather than a comparative approach, this framework takes a predictive
approach to inference \cite{Breiman2001-uz}. We use a linear modeling approach,
which aims to predict phenotypical variance in a group of subjects, based on a
linear combination of the features estimated with tractometry. Using this
approach, we need to first deal with the large and asymmetric dimensionality of
the data: tractometry data usually has many more features (i.e., number of
measurements per individual) than samples (number of subjects), which makes
inferences from the data about phenotypical differences between individuals
ill-posed. This regime is the target of several statistical learning techniques,
and is often solved by various forms of regularization. For example, Tikhonov
regularization shrinks the solution such that the sum of squared contributions
from the individual features are minimized \cite{Hoerl2000-ij}. Another solution
to the problem is provided by the Lasso algorithm, which instead minimizes the
sum of the absolute values of contributions of each feature
\cite{Tibshirani1996-qs}. This tends to shrink to zero the contributions of many
of the features, providing results that are both accurate and interpretable.
When additional structure is available in the organization of the data,
regularization algorithms can take advantage of this structure. For example, if
the features lend themselves to a natural division into different groups, the
group lasso (GL) can be used to select groups of features, rather than
individual features \cite{Yuan2006-ky}. The Sparse Group Lasso (SGL) elaborates
on this idea by providing control both of group sparsity, as well as overall
sparsity of the solutions \cite{simon2013sgl}. Because the features measured
with tractomery lend themselves to grouping based on the tracts from which each
measurement is taken, GL and SGL provide a useful tool linear model fitting in
problems of this form. Using linear regression as the basis for analysis is
useful for both regression tasks, in which a continuous measure (e.g., age) is
modeled, as well as for classification tasks, where a discrete property is
modeled (e.g., whether an individual has a particular disease). Moreover, use of
SGL helps derive informative properties of the human white matter, both in cases
where the underlying causes of variance are sparsely concentrated (e.g., in one
particular white matter tract) and in cases where biological substrates of
differences between individuals are more widespread. This is because SGL is
controlled by meta-parameters that set the degree to which group sparsity and
overall sparsity are applied. These metaparameters can be set empirically, based
on the data, through a process of nested cross-validation. We demonstrate these
properties of the algorithm in data from previously published studies
\cite{sarica2017corticospinal, yeatman2014lifespan}.

### 2019-12-06 reports/manuscript/sections/methods.tex

\item Diffusion MRI data from a previous study of properties of the white matter
across the lifespan\cite{yeatman2014lifespan}, containing dMRI data from 76
subjects with ages 6-50. These data were measured in a GE Discovery 750 3T MRI
scanner at the Stanford Center for Cognitive and Neurobiological Imaging. The
Stanford University IRB approved the procedures of this study. Informed consent
was obtained from each adult participant, and assent for participation was
provided by parents/guardians for children. Voxel resolution was
\num{2x2x2}$mm^3$ with 96 non-colinear directions measured with a $b=2000$
$\frac{sec}{mm^2}$ and 30 non-colinear directions measured with a $b=1000$
$\frac{sec}{mm^2}$. Data was preprocessed to correct for subject motion and the
diffusion tensor model \cite{basser1994mr} was fit in every voxel, using a
robust fit \cite{chang2005restore}. These data were acquired using a dual spin
echo sequence, in which there is sufficient time for eddy currents to subside
between the application of the gradients and the image acquisition, so no eddy
current correction was appliced. We will refer to this dataset as WH.

\item Diffusion MRI from from a previous study of the corticospinal tract
profile in patients with amyotrophic lateral sclerosis
(ALS)\cite{sarica2017corticospinal}, containing data from 24 ALS patients and 24
demographically matched healthy controls. These data were measured in a GE
Discovery 750 3T MRI scanner at the Institute of Bioimaging and Molecular
Physiology in Catanzaro. Informed consent was provided as approved by the
Ethical Committee of the University ``Magna Graecia'' of Catanzaro. Voxel
resolution was \num{2x2x2} $mm^3$ and 27 non-colinear directions were
measured with a $b=1000$ $\frac{sec}{mm^2}$. Here, both motion correction and
eddy current correction were applied before modeling the diffusioon tensor in
each voxel. We will refer to this dataset as ALS.

@@

Here, we propose that in addition to regularizing the coefficients in $\beta$,
we can also capitalize on our knowledge of the group structure of the tract
profile features in $\mathbf{X}$. The tract-metric combinations form a natural
grouping. For example, we expect that MD features within the left arcuate
fasciculus will co-vary across individuals. Likewise for FA values within the
right corticospinal tract (CST) and so on. This group structure is represented
in Fig~\ref{fig:group-structure}, which depicts the linear model $y = \mathbf{X}
\cdot \beta$. Thus, we seek a regularization approach that will fit a linear
model with anatomically-grouped covariates, where we expect to observe both
groupwise sparsity, where the number of groups (tract/metric combinations) with
at least one non-zero coefficients is small, as well as within-group sparsity,
where the number of non-zero coefficients within each non-zero group is small.
The sparse group lasso (SGL)
is a penalized regression technique that satisfies exactly these
criteria\cite{simon2013sparse}. It solves for a coefficient vector
$\hat{\beta}$ that satisfies
\begin{equation}
    \hat{\beta} = \min_\beta \frac{1}{2}
    ||y - \displaystyle \sum_{\ell = 1}^{G}
    \mathbf{X}^{(\ell)} \cdot \beta^{(\ell)}||_2^2
    + \lambda_1 \displaystyle \sum_{\ell = 1}^{G}
    \sqrt{p_\ell} ||\beta^{(\ell)}||_2
    + \lambda_2 ||\beta||_1,
    \label{eq:sgl}
\end{equation}

@@

For given values of $\lambda_1$ and $\lambda_2$, the cost function in equation
\ref{eq:sgl} can be optimized using proximal gradient descent methods
\cite{parikh2014proximal} here implemented as a custom proximal operator that is
then optimized using the C-OPT library\cite{copt}. This supplies an estimate of
the optimal $\hat{\beta}$ given a particular set of values for the
meta-parameters $\lambda_1$ and $\lambda_2$.

To objectively evaluate the model and guard against over-fitting, we used a
nested cross-validation scheme, depicted in Fig~\ref{fig:cross-val} for the
categorical classification case. The input data is shuffled and then decomposed
into $k$ batches, hereafter referred to as folds. For the ALS dataset we used
$k=10$ and for the WH dataset $k=5$. For each unique fold, we hold that fold out
as the test\textsubscript{outer} set and let the remaining data comprise the
train\textsubscript{outer} set, with the subscript indicating the depth of the
nested decomposition. We further decompose each train\textsubscript{outer} set
into three folds, and again for each unique fold, we hold out that fold as the
test\textsubscript{inner} set and let the remaining data comprise the
train\textsubscript{inner} set. At level-1 of the decomposition, we fit an SGL
model using fixed regularization meta-parameters $\lambda_1$ and $\lambda_2$,
training the model using train\textsubscript{inner} and evaluating the fit on
test\textsubscript{inner}. We find the optimal values for $\lambda_1$ and
$\lambda_2$ using hyperoptimizatin. This is implemented using the hyperopt
library's \verb|fmin| function\cite{Bergstra_2015} with a tree of Parzen
estimators search algorithm\cite{bergstra2011algorithms}. For continuous
numerical $y$, \verb|fmin| searches for meta-parameter values that minimize the
median absolute error. This can also be done minimizing the root of the mean
squared error (RMSE) or to maximizing the coefficient of determination ($R^2$).
For binary categorical $y$ \verb|fmin| seeks to maximize the classification
accuracy. This can also be done maximizing the area under the receiver operating
curve (ROC AUC) or the average precision. Using hyperoptimization, we find
optimal regularization parameters and $\hat{\beta}$ for each
train\textsubscript{outer} set and then use those to predict values for data in
test\textsubscript{outer}. Thus each subject in the dataset has a predicted
phenotype derived from a model that never saw its particular subject's data.

>>>>

\item Diffusion MRI data from a previous study of properties of the white matter
across the lifespan\cite{yeatman2014lifespan}, containing dMRI data from 76
subjects with ages 6-50. These data were measured in a GE Discovery 750 3T MRI
scanner at the Stanford Center for Cognitive and Neurobiological Imaging. The
Stanford University IRB approved the procedures of this study. Informed consent
was obtained from each adult participant, and assent for participation was
provided by parents/guardians for children. Voxel resolution was
\num{2x2x2}\si{\milli\meter^3} with 96 non-colinear directions measured with a
$b= \SI{2000}{\second\per\milli\meter^2}$ and 30 non-colinear directions
measured with a $b=\SI{1000}{\second\per\milli\meter^2}$.
Data was preprocessed to correct for subject motion and the diffusion
tensor model \cite{basser1994mr} was fit in every voxel, using a robust fit
\cite{chang2005restore}. These data were acquired using a dual spin echo
sequence, in which there is sufficient time for eddy currents to subside between
the application of the gradients and the image acquisition, so no eddy current
correction was appliced. We will refer to this dataset as WH.

\item Diffusion MRI from from a previous study of the corticospinal tract
profile in patients with amyotrophic lateral sclerosis
(ALS)\cite{sarica2017corticospinal}, containing data from 24 ALS patients and 24
demographically matched healthy controls. These data were measured in a GE
Discovery 750 3T MRI scanner at the Institute of Bioimaging and Molecular
Physiology in Catanzaro. Informed consent was provided as approved by the
Ethical Committee of the University ``Magna Graecia'' of Catanzaro. Voxel
resolution was \num{2x2x2}\si{\milli\meter^3} and 27 non-colinear directions
were measured with a $b=\SI{1000}{\second\per\milli\meter^2}$.
Here, both motion correction and eddy current correction were
applied before modeling the diffusioon tensor in each voxel. We will refer to
this dataset as ALS.

@@

Here, we propose that in addition to regularizing the coefficients in $\beta$,
we can also capitalize on our knowledge of the group structure of the tract
profile features in $\mathbf{X}$. The tract-metric combinations form a natural
grouping. For example, we expect that MD features within the left arcuate
fasciculus will co-vary across individuals. Likewise for FA values within the
right corticospinal tract (CST) and so on. This group structure is represented
in Fig~\ref{fig:group-structure}, which depicts the linear model $y = \mathbf{X}
\cdot \beta$. Thus, we seek a regularization approach that will fit a linear
model with anatomically-grouped covariates, where we expect to observe both
groupwise sparsity, where the number of groups (tract/metric combinations) with
at least one non-zero coefficients is small, as well as within-group sparsity,
where the number of non-zero coefficients within each non-zero group is small.
The sparse group lasso (SGL)
is a penalized regression technique that satisfies exactly these
criteria\cite{simon2013sparse}. It solves for a coefficient vector
$\widehat{\beta}$ that satisfies
\begin{equation}
    \widehat{\beta} = \min_\beta \frac{1}{2}
    \norm*{y - \displaystyle \sum_{\ell = 1}^{G}
    \mathbf{X}^{(\ell)} \cdot \beta^{(\ell)}}_2^2
    + \lambda_1 \displaystyle \sum_{\ell = 1}^{G}
    \sqrt{p_\ell} \norm*{\beta^{(\ell)}}_2
    + \lambda_2 \norm*{\beta}_1,
    \label{eq:sgl}
\end{equation}

@@

For given values of $\lambda_1$ and $\lambda_2$, the cost function in equation
\ref{eq:sgl} can be optimized using proximal gradient descent methods
\cite{parikh2014proximal} here implemented as a custom proximal operator that is
then optimized using the C-OPT library\cite{copt}. This supplies an estimate of
the optimal $\widehat{\beta}$ given a particular set of values for the
meta-parameters $\lambda_1$ and $\lambda_2$.

To objectively evaluate the model and guard against over-fitting, we used a
nested cross-validation scheme, depicted in Fig~\ref{fig:cross-val} for the
categorical classification case. The input data is shuffled and then decomposed
into $k$ batches, hereafter referred to as folds. For the ALS dataset we used
$k=10$ and for the WH dataset $k=5$. For each unique fold, we hold that fold out
as the test\textsubscript{outer} set and let the remaining data comprise the
train\textsubscript{outer} set, with the subscript indicating the depth of the
nested decomposition. We further decompose each train\textsubscript{outer} set
into three folds, and again for each unique fold, we hold out that fold as the
test\textsubscript{inner} set and let the remaining data comprise the
train\textsubscript{inner} set. At level-1 of the decomposition, we fit an SGL
model using fixed regularization meta-parameters $\lambda_1$ and $\lambda_2$,
training the model using train\textsubscript{inner} and evaluating the fit on
test\textsubscript{inner}. We find the optimal values for $\lambda_1$ and
$\lambda_2$ using hyperoptimizatin. This is implemented using the hyperopt
library's \verb|fmin| function\cite{Bergstra_2015} with a tree of Parzen
estimators search algorithm\cite{bergstra2011algorithms}. For continuous
numerical $y$, \verb|fmin| searches for meta-parameter values that minimize the
median absolute error. This can also be done minimizing the root of the mean
squared error (RMSE) or to maximizing the coefficient of determination ($R^2$).
For binary categorical $y$ \verb|fmin| seeks to maximize the classification
accuracy. This can also be done maximizing the area under the receiver operating
curve (ROC AUC) or the average precision. Using hyperoptimization, we find
optimal regularization parameters and $\widehat{\beta}$ for each
train\textsubscript{outer} set and then use those to predict values for data in
test\textsubscript{outer}. Thus each subject in the dataset has a predicted
phenotype derived from a model that never saw its particular subject's data.
